Binning Strategies and Related Loss for Large Data

### 2015-03-19 Paper.Rnw
move old "information recovery" subsection into new subsection in binning section. Discuss binned data reduction (one row per point -> one row per bin with freq column added).  Reduced binned data is concept necessary to be referenced in loss section.

@@

\subsection{Recovery of Numerical Information}
% The total loss from reducing a value $x$ to bin center $x^\ast$ is at least as large as the loss from reducing points $x_i$ to $\overline x_i$, the numerical center of the binned data.
% % that's only true for the expected value - not for all individuals, but you don't need - we just use the partition.

As shown before, the overall loss partitions naturally into visual and numerical loss (see equation \hh{XXX need new equation reference} %\ref{eqn:partition}
). We can make use of this and
 recover some of the numerical information lost due to binning by storing the numerical center $(\bar{x},\bar{y})$ instead of the visual center $(x^\ast,y^\ast)$ for use in subsequent calculations. Storing the binned data as the arithmetic mean rather than the visual center reduces the total loss, and the visual center can be recovered from the bin width and  bin center $(O_X, O_Y)$, see table \ref{RecoveryComparison} for a numeric example for this.

<<recover,echo=FALSE,eval=FALSE,results='asis'>>=
set.seed(90)
x <- rnorm(100, 5, 2)
y <- rnorm(100, 5, 2)
d <- data.frame(x=x, y=y, c=1)
d <- d[order(d$x, d$y),]
print(xtable(head(d,10)[c(2,1,3,6,10,5,4,8,9,7),], digits=c(0, 4, 4, 0), caption=paste('Original Data\\newline', nrow(d), 'rows'), include.rownames=FALSE, table.placement="hbtp", floating=FALSE))
dbin <- ddply(binStd(d, c(1,1)), .(x,y), summarize, c=sum(c))
dbin <- dbin[order(dbin$x, dbin$y),]
print(xtable(head(dbin), digits=c(0, 0, 0, 0), caption=paste('Binned Data\\newline', nrow(dbin), 'rows'), align=c('X','X','X','c')), include.rownames=FALSE, table.placement="hbtp", floating=FALSE, tabular.environment="tabularx", width='.7\\textwidth')
names(d) <- c("x", "y", "Freq")
drecover <- unique(lossCalc(d, c(1,1), "standard", newData=TRUE)$NewData)
names(drecover) <- c("x", "y", "c")
drecover <- drecover[order(drecover$x, drecover$y),]
print(xtable(head(drecover), digits=c(0, 4, 4, 0), caption=paste('Recovered Data\\newline', nrow(drecover), 'rows'), include.rownames=FALSE, table.placement="hbtp", floating=FALSE))
@

\begin{table}[hbtp]
\begin{minipage}[t]{.33\textwidth}\centering\subfloat[Original Data, 100 rows]{
\begin{tabular}{rrr}\hline
       $x$ &      $y$ & c \\ \hline
 -0.2582 & 1.5497 & 1 \\ 
 -0.3843 & 5.6508 & 1 \\ 
 1.0597 & 2.4308 & 1 \\ 
 2.2680 & 1.5877 & 1 \\ 
 2.4992 & 4.6260 & 1 \\ 
 1.6788 & 5.0260 & 1 \\ 
 1.6054 & 6.4125 & 1 \\ 
 2.4180 & 6.4438 & 1 \\ 
 2.4345 & 6.4462 & 1 \\ 
 2.3620 & 10.3119 & 1 \\  \hline
\end{tabular}}
\end{minipage}\hfil
\begin{minipage}[t]{.33\textwidth}\centering\subfloat[Binned Data Visual Centers, 49 rows]{
\begin{tabularx}{.7\textwidth}{XXc} \hline
  $x^\ast$ & $y^\ast$ & c \\ \hline
  0 & 2 & 1 \\ 
  0 & 6 & 1 \\ 
  1 & 2 & 1 \\
  2 & 2 & 1 \\
  2 & 5 & 2 \\ 
  2 & 6 & 3 \\ 
  2 & 10 & 1 \\ 
   \hline
\end{tabularx}}
\end{minipage}\hfil
\begin{minipage}[t]{.33\textwidth}\centering\subfloat[Binned Data Numerical Centers, 49 rows]{
\begin{tabular}{rrc}\hline
  $\bar{x}$   &    $\bar{y}$ & c \\ \hline
 -0.3843 & 5.6508 & 1 \\ 
 -0.2582 & 1.5497 & 1 \\ 
  1.0597 & 2.4308 & 1 \\ 
  2.2680 & 1.5877 & 1 \\  
  2.0890 & 4.8260 & 2 \\ 
  2.1526 & 6.4341 & 3 \\ 
  2.2680 & 1.5877 & 1 \\ \hline
\end{tabular}}
\end{minipage}
\caption{\label{RecoveryComparison}ten rows from Original and Binned Data Tables, with data storage sizes}
\end{table}

In spite of the costs, information recovery is still useful for very large data, where the unbinned data set is  computationally intractable. Minimizing numerical loss in this manner provides for more accurate calculations from the data, while maintaining the storage space advantages of binning. %but creates an additional complexity in that the geometric center of the tile (the visual bin center) does not correspond to the stored, binned data. 
% The random binning algorithm serves as a less computationally intensive, unbiased alternative for 
% data sets which are too large to partition the loss in this manner.
% \hh{is that true, or does it require the expected value to be the visual center?}

>>>>

\subsection{Binned Data Reduction}

\ktm{Section to describe how data is stored post binning. Section will discuss difference between binned data (bin centers added to original data, still n rows) and "reduced binned data" (aggregated into frequencies), sufficiency for constructing binned scatterplot and storage size.  Want to reference these ideas in later sections.}

% As shown before, the overall loss partitions naturally into visual and numerical loss (see equation \hh{XXX need new equation reference} %\ref{eqn:partition}
% ). We can make use of this and
%  recover some of the numerical information lost due to binning by storing the numerical center $(\bar{x},\bar{y})$ instead of the visual center $(x^\ast,y^\ast)$ for use in subsequent calculations. Storing the binned data as the arithmetic mean rather than the visual center reduces the total loss, and the visual center can be recovered from the bin width and  bin center $(O_X, O_Y)$, see table \ref{RecoveryComparison} for a numeric example for this.

<<DataReduction,echo=FALSE,eval=FALSE,results='asis'>>=
n=12
param1 = 1
param2 = 1.6
set.seed(52)
x <- rbeta(n,param1, param2)*20-10
y <- rbeta(n,param1, param2)*20-10
# standard bin centers
xcent <- rep(-5,n) ; xcent[x>0] <- 5
ycent <- rep(-5,n) ; ycent[y>0] <- 5
idx <- order(xcent,ycent)
d <- data.frame(x,y,xcent,ycent)[idx,]
dbin <- ddply(d, .(xcent,ycent), summarise,
              c=length(x))
dbin <- dbin[order(dbin$xcent,dbin$ycent),]

print(xtable(d[1:2], digits=c(0, 4, 4), caption=paste('Original Data\\newline', nrow(d), 'rows')), include.rownames=FALSE, table.placement="hbtp", floating=FALSE)

print(xtable(d[3:4], digits=c(0, 0, 0), caption=paste('Binned Data\\newline', nrow(d), 'rows')), include.rownames=FALSE, table.placement="hbtp", floating=FALSE)

print(xtable(dbin, digits=c(0, 0, 0, 0), caption=paste('Binned Data\\newline', nrow(dbin), 'rows'), align=c('X','X','X','c')), include.rownames=FALSE, table.placement="hbtp", floating=FALSE, tabular.environment="tabularx", width='.7\\textwidth')

@

\begin{table}[H]
\begin{minipage}[t]{.33\textwidth}\centering\subfloat[Original Data, 12 rows]{
\begin{tabular}{rr}\hline
       $x$ &  $y$  \\ \hline
-7.7325 & -9.6340 \\ 
  -8.1176 & -1.4529 \\ 
  -5.8996 & -3.2033 \\ 
  -7.0375 & -5.5563 \\ 
  -3.6354 & -3.9315 \\ 
  -8.7639 & 0.9874 \\ 
  -2.9781 & 8.6802 \\ 
  0.8210 & -8.6118 \\ 
  5.4477 & -8.4555 \\ 
  4.6849 & -5.6620 \\ 
  9.4785 & 1.1133 \\ 
  1.7579 & 5.3759 \\  \hline
\end{tabular}}
\end{minipage}\hfil
\begin{minipage}[t]{.33\textwidth}\centering\subfloat[Binned Data Centers, 12 rows]{
\begin{tabularx}{.7\textwidth}{rr} \hline
  $b_X(x)$ & \hspace{.4cm} $b_Y(y)$  \\ \hline
  -5 & -5 \\ 
  -5 & -5 \\ 
  -5 & -5 \\ 
  -5 & -5 \\ 
  -5 & -5 \\ 
  -5 & 5 \\ 
  -5 & 5 \\ 
  5 & -5 \\ 
  5 & -5 \\ 
  5 & -5 \\ 
  5 & 5 \\ 
  5 & 5 \\  \hline
\end{tabularx}}
\end{minipage}\hfil
\begin{minipage}[t]{.33\textwidth}\centering\subfloat[Reduced Binned Data, 4 rows]{
\begin{tabularx}{.7\textwidth}{rrr}\hline
 \hspace{.2cm} $x^\ast$   &   \hspace{.2cm} $y^\ast$ & \hspace{.4cm}c \\ \hline
-5 & -5 & 5 \\ 
  -5 & 5 & 2 \\ 
  5 & -5 & 3 \\ 
  5 & 5 & 2 \\  \hline
\end{tabularx}}
\end{minipage}
\label{DataReductionExample}
\caption{Original, Binned and Reduced Binned Data Tables, with data storage sizes.  Binned using standard rectangular approach with origin at (-10,-10) and binwidths $\omega_x = \omega_y = 10$. }
\end{table}

In spite of the costs, information recovery is still useful for very large data, where the unbinned data set is  computationally intractable. Minimizing numerical loss in this manner provides for more accurate calculations from the data, while maintaining the storage space advantages of binning.

@@

%-----------------------------------------------------------------------------
